# Aside: Animate a NN

```{r setup, include=FALSE}
set.seed(155) #151
```

## Libraries
```{r libs,warning=FALSE}
library(tidyverse)
library(nnet)
library(ggthemes)
library(gganimate)
library(av)

```

## The Data
Here is a quick demo of how a neural network successively iterates to improve fit. We will create a simple sine wave and then ask a neural net to try to predict it. We will animate that process by saving the prediction at each iteration. 

Here is the function and the data to fit.

```{r the function}
n <- 400
dat <- tibble(x = seq(0,n),
              y = sin(2 * pi / 100 * seq(0,n)))
dat %>% ggplot(mapping = aes(x=x,y=y)) +
  geom_line(color="#fbe4c6", alpha=0.7) +
  labs(x=element_blank(),y=element_blank()) + 
  theme_hc(style = 'darkunica') +
  theme(legend.position = "none",
        text=element_text(size=12))
```

## Predicting

We will use a single-hidden-layer neural network to fit to the data. We do some cumbersome looping in order to save the predictions from each iteration. If the fit is really really good (r > 0.999) the loop will quit. With this simple sine function, that criteria is usually met.

```{r run nnet,warning=FALSE}
nIter <- 100
fits <- as_tibble(matrix(0,nrow=length(dat$x),ncol=nIter))
gof <- tibble(Iter = 1:nIter, Correlation = numeric(nIter), RMSE = numeric(nIter)) 
for(i in 1:nIter){
  if(i==1) {
    # because i=1 we start without weights
    nn <- nnet(dat$x, dat$y, size=10, maxit=i-1, linout=TRUE, trace = FALSE)
  }
  else {
    # use prior weights to start the nnet once they exist (i>1)
    nn <- nnet(dat$x, dat$y, size=10, maxit=i-1, linout=TRUE, trace = FALSE, Wts = nn$wts)
  }
  yhat <- predict(nn)[,1]
  fits[,i] <- yhat
  gof$Correlation[i] <- cor(dat$y,yhat)
  gof$RMSE[i] <- sqrt(mean((dat$y-yhat)^2))
  # if fits are ~perfect, stop
  if(gof$Correlation[i] > 0.9995) { break }
}
# trim fits if loop stopped early
fits <- fits[,1:i]
fits2 <- as_tibble(cbind(dat$x,fits)) 
names(fits2)[1] <- "x"
names(fits2)[-1] <- formatC(1:i,digits = 1,flag=0)
gof <- gof[1:i,]
tail(gof)
```

Above, the loop reached `r i` iterations before the fit was essentially perfect. Let's see all the predictions.

```{r plot,warning=FALSE}
fits2 <- fits2 %>% 
  pivot_longer(cols=-1,names_to = "IterChar", values_to = "yhat") %>%
  mutate(Iter = as.numeric(IterChar))

p1 <- ggplot(fits2,aes(x=x,y=yhat,color=IterChar)) +
  # original data
  geom_line(data=dat,aes(x=x,y=y),inherit.aes = FALSE, 
            color="#fbe4c6", alpha=0.7) +
  # nn fits
  geom_line() + 
  scale_color_manual(values=PNWColors::pnw_palette("Spring",i)) +
  labs(x=NULL,y=NULL) +
  theme_hc(style = 'darkunica') +
  theme(legend.position = "none",
        text=element_text(size=12))
p1
```

And the evolution of the RMSE and the correlation (r).

```{r get gof}
gof2 <- gof %>% pivot_longer(cols = -1)    
p2 <- ggplot(gof2,aes(x=Iter,y=value,color=Iter)) +
  geom_point() + geom_line() +
  scale_color_gradientn(colors=PNWColors::pnw_palette("Spring",i)) +
  labs(x="Iteration",y=NULL) +
  facet_wrap(~name) +
  theme_hc(style = 'darkunica') +
  theme(legend.position = "none",
        text=element_text(size=12))
p2
```

## Animate
And now we can animate it all.

```{r first anim, eval = TRUE}
anim1 <- p1 + 
  labs(title = 'Iteration: {floor(frame_time)}') +
  transition_time(Iter)

animate(anim1, duration = i, detail = 1, 
        height = 5, width = 5, units = "in", res=250,
        renderer = ffmpeg_renderer(format = "webm"))

```


```{r save gif, echo=FALSE, eval=TRUE}
anim1_gif <- animate(anim1, 
                     height = 5, width = 5, units = "in", res=250,
                     renderer = magick_renderer(),
                     start_pause=10, end_pause = 10)
anim_save("nnFits_anim.gif",animation=anim1_gif)
```

And below is a combo of the fits and the skill in a GIF. Note that combining the two plots to have it all work side by side is a little more complicated than I'd like it to be and we have to use the `magick` library. See [here](https://github.com/thomasp85/gganimate/wiki/Animation-Composition). I hid the code because it's yucky and confusing.

```{r make gifs and append, eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}
library(magick)

anim2 <- p2 + labs(x = NULL) + 
  transition_reveal(Iter)

# changed renderer to improve quality
anim1_gif <- animate(anim1,
                     height = 5, width = 5, units = "in", res=250,
                     renderer = magick_renderer())
anim2_gif <- animate(anim2,
                     height = 5, width = 5, units = "in", res=250,
                     renderer = magick_renderer()) 

combined_gif <- image_append(c(anim1_gif[1], anim2_gif[1]))
nFrames <- dim(image_info(anim1_gif))[1]
for(j in 2:nFrames){
  tmp <- image_append(c(anim1_gif[j], anim2_gif[j]))
  combined_gif <- c(combined_gif, tmp)
}
anim_save("nnPreds_Skill_Anim.gif",animation=combined_gif)
combined_gif
```


